Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-27 articles/cycle/alps-cycle.tex
Implement comments from Guillaume.

    Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
    glacial cycle (120--0\,ka) in the Alps, using horizontal resolutions of 2
    and 1\,km. Climate forcing is derived using present-day climate data from
    WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
    offsets from multiple palaeo-climate proxies, among which only the EPICA ice
    core record yields glaciation during marine oxygen isotope stages~4
    (69--62\,ka) and~2 (34--18\,ka) spatially and temporally consistent with
    the geological reconstructions, while the other records used result in
    excessive early glacial cycle ice cover and a late Last Glacial Maximum.
    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice sheet, showing that Alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences through the
    mountain passes. Modelled maximum ice surface is 861\,m
    higher than observed trimline elevations in the upper Rhone Valley, yet
    our simulation predicts little erosion at high elevation due to cold-based
    ice. Finally, despite the uniform climate forcing, differences in glacier
    catchment hypsometry cause the Last Glacial Maximum advance to be modelled
    as a time-transgressive event, with some glaciers reaching their maximum as
    early as 27\,ka, and others as late as 21\,ka.

@@

    The spatial extent and thickness of Alpine glaciers during the LGM, an
    integrated footprint of thousands of years of climate history and glacier
    dynamics, have been reconstructed from moraines and trimlines across the
    mountain range \citep[e.g.,][]{Penck.Bruckner.1909, Castiglioni.1940,
    Husen.1987, Bini.etal.2009, Coutterand.2010}. The assumption that
    trimlines, the upper limit of glacial
    erosion, represent the maximum ice surface elevation, has
    been repeatedly invalidated in other glaciated regions of the globe
    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012,
    Ballantyne.Stone.2015}. Although
    no evidence for cold-based glaciation has been reported in the Alps to
    date, it is supported by numerical modelling
    \citep[Fig.~3]{Haeberli.Schluchter.1987}. Alpine ice flow
    patterns were primarily controlled by subglacial topography, but there is
    evidence for flow accross major mountain passes
    \citep[e.g.,][]{Coutterand.2010, Kelly.etal.2004, Husen.2011}, and
    self-sustained ice domes \citep{Bini.etal.2009}. Finally, the timing of the
    LGM in the Alps \citep{Ivy-Ochs.etal.2008, Monegato.etal.2017} is in good
    agreement with the maximum expansion of continental ice sheets recorded by
    the Marine Oxygen Isotope Stage (MIS) 2
    \citep[29--14\,ka;][]{Lisiecki.Raymo.2005}. Regional variation between
    different piedmont
    lobes exists \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear whether
    this relates to climate or glacier dynamics \citep{Monegato.etal.2017},
    uncertainties in the dating methods, or both.

@@

    We use the Parallel Ice Sheet Model (PISM, development version~e9d2d1f), an
    open source, finite difference, shallow ice sheet model
    \citep{PISM-authors.2017}. The model requires input on initial bedrock and
    glacier topographies, geothermal heat flux and climate forcing. It computes
    the
    evolution of ice extent and thickness over time, the thermal and dynamic
    states of the ice sheet, and the associated lithospheric response. The
    model set-up used here was previously developed and tested on the former
    Cordilleran ice sheet \citep{Seguinot.2014, Seguinot.etal.2014,
    Seguinot.etal.2016} and subsequently adapted for steady climate
    \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a}
    applications to the European Alps.

@@

    Ice sheet dynamics are typically modelled using a~combination of internal
    deformation and basal sliding. PISM is a~shallow ice sheet model, which
    implies that the balance of stresses is approximated based on their
    predominant components. The Shallow Shelf Approximation (SSA) is combined
    with the Shallow Ice Approximation (SIA) by adding velocity solutions of
    the two approximations \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}.
    Although this heuristic approach is inaccurate in the transition zone
    where both vertical shear and longitudinal stretch are significant,
    it was shown to effectively approximate complex ice sheet flow
    dynamics over mountainous topographies similar to that of the European Alps
    \citep{Golledge.etal.2012, Ziemen.etal.2016}. This hybrid scheme is
    computationally much more efficient than a fully three-dimensional model
    \citep[e.g.,][]{Cohen.etal.2017}
    with which simulations of the scale presented here would not be feasible.

@@

    \begin{figure}[t]
      \centerline{\includegraphics{alpcyc_hr_trimlines}}
      \caption{%
        \textbf{(a)} Comparison of modelled ice surface elevation at the LGM
        (time-transgressive, corresponding to maximum ice thickness,
        Fig.~\ref{fig:timing}), compensated for bedrock deformation, against
        observed trimline elevations for the upper Rhone Glacier
        \citep[Table~1]{Kelly.etal.2004}. Colours
        show the age of maximum ice thickness (as in Fig.~\ref{fig:timing}).
        Model variables were bilinearly interpolated to the trimline locations.
        \textbf{(b)} Histogram of differences between modelled LGM ice surface
        elevation and trimline elevations (50\,m bands). The average
        difference is 861\,m.
        \textbf{(c)} Observed upper Rhone Valley trimline locations
        \citep[Table~1]{Kelly.etal.2004}, modelled age of maximum ice thickness
        (colour) and LGM ice surface elevation (200\,m contours). Hatches mark
        the LGM cold-based ares (basal temperature above $1 \times 10^{-3}$\,K
        below freezing at the age of maximum ice thickness).}
      \label{fig:trimlines}
    \end{figure}

@@

    In the upper Rhone Valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of $861\pm197$\,m (one standard deviation,
    Fig.~\ref{fig:trimlines}b), which is significant in respect to the modelled
    maximum ice thickness in the Rhone Valley (and the Alps) of 2586\,m. This
    result depends on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here. However, a previous
    sensitivity study conducted with PISM and a very similar model set-up on
    the Cordilleran ice sheet show that varying ice rheological parameters
    resulted in relative errors in modelled total ice volume of 25\,\unit{\%}
    with regard to ice rheological parameters and 14\,\unit{\%} regarding basal
    sliding parameters \citep[Fig.~7]{Seguinot.etal.2016}. In addition,
    regional sensitivity tests in the Rhone Valley show that a modelled ice
    surface compatible with the trimline elevations is incompatible with the
    documented overspilling flow to the south across the Simplon pass
    \citep{Becker.etal.2017}.

>>>>

    Here, we use the Parallel Ice Sheet Model (PISM) to model the entire last
    glacial cycle (120--0\,ka) in the Alps, using horizontal resolutions of 2
    and 1\,km. Climate forcing is derived using present-day climate data from
    WorldClim and the ERA-Interim reanalysis, and time-dependent temperature
    offsets from multiple palaeo-climate proxies, among which only the EPICA ice
    core record yields glaciation during marine oxygen isotope stages~4
    (69--62\,ka) and~2 (34--18\,ka) spatially and temporally consistent with
    the geological reconstructions, while the other records used result in
    excessive early glacial cycle ice cover and a late Last Glacial Maximum.
    Despite the low variability of this Antarctic-based climate forcing, our
    simulation depicts a highly dynamic ice sheet, showing that Alpine glaciers
    may have advanced many times over the foreland during the last glacial
    cycle. Ice flow patterns during peak glaciation are largely governed by
    subglacial topography but include occasional transfluences through the
    mountain passes. Modelled maximum ice surface is in average 861\,m
    higher than observed trimline elevations in the upper Rhone Valley, yet
    our simulation predicts little erosion at high elevation due to cold-based
    ice. Finally, despite the uniform climate forcing, differences in glacier
    catchment hypsometry cause the Last Glacial Maximum advance to be modelled
    as a time-transgressive event, with some glaciers reaching their maximum as
    early as 27\,ka, and others as late as 21\,ka.

@@

    The spatial extent and thickness of Alpine glaciers during the LGM, an
    integrated footprint of thousands of years of climate history and glacier
    dynamics, have been reconstructed from moraines and trimlines across the
    mountain range \citep[e.g.,][]{Penck.Bruckner.1909, Castiglioni.1940,
    Husen.1987, Bini.etal.2009, Coutterand.2010}. The assumption that
    trimlines, the upper limit of glacial
    erosion, represent the maximum ice surface elevation, has
    been repeatedly invalidated in other glaciated regions of the globe
    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012,
    Ballantyne.Stone.2015}. Although
    no evidence for cold-based glaciation has been reported in the Alps to
    date, it is supported by numerical modelling
    \citep[Fig.~3]{Haeberli.Schluchter.1987, Cohen.2017}. Alpine ice flow
    patterns were primarily controlled by subglacial topography, but there is
    evidence for flow accross major mountain passes
    \citep[e.g.,][]{Coutterand.2010, Kelly.etal.2004, Husen.2011}, and
    self-sustained ice domes \citep{Bini.etal.2009}. Finally, the timing of the
    LGM in the Alps \citep{Ivy-Ochs.etal.2008, Monegato.etal.2017} is in good
    agreement with the maximum expansion of continental ice sheets recorded by
    the Marine Oxygen Isotope Stage (MIS) 2
    \citep[29--14\,ka;][]{Lisiecki.Raymo.2005}. Regional variation between
    different piedmont
    lobes exists \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear whether
    this relates to climate or glacier dynamics \citep{Monegato.etal.2017},
    uncertainties in the dating methods, or both.

@@

    We use the Parallel Ice Sheet Model (PISM, development version~e9d2d1f), an
    open source, finite difference, shallow ice sheet model
    \citep{PISM-authors.2017}. The model requires input on initial bedrock and
    glacier topographies, geothermal heat flux and climate forcing. It computes
    the
    evolution of ice extent and thickness over time, the thermal and dynamic
    states of the ice sheet, and the associated lithospheric response. The
    model set-up used here was previously developed and tested on the former
    Cordilleran ice sheet \citep{Seguinot.2014, Seguinot.etal.2014,
    Seguinot.etal.2016} and subsequently adapted for steady climate
    \citep{Becker.etal.2016} and regional \citep{Jouvet.etal.2017a,
    Becker.Becker.etal.2017} applications to the European Alps.

@@

    Ice sheet dynamics are typically modelled using a~combination of internal
    deformation and basal sliding. PISM is a~shallow ice sheet model, which
    implies that the balance of stresses is approximated based on their
    predominant components. The Shallow Shelf Approximation (SSA) is combined
    with the Shallow Ice Approximation (SIA) by adding velocity solutions of
    the two approximations \citep[Eqs.~7--9 and 15]{Winkelmann.etal.2011}.
    Although this heuristic approach is inaccurate in the transition zone
    where both vertical shear and longitudinal stretch are significant,
    it was shown to effectively approximate complex ice sheet flow
    dynamics over mountainous topographies similar to that of the European Alps
    \citep{Golledge.etal.2012, Ziemen.etal.2016}. In the SIA, topographic
    roughness using a bed smoother range of 5\,km \citep{Schoof.2003}.

@@

    \begin{figure}[t]
      \centerline{\includegraphics{alpcyc_hr_trimlines}}
      \caption{%
        \textbf{(a)} Comparison of modelled ice surface elevation at the LGM
        (time-transgressive, corresponding to maximum ice thickness,
        Fig.~\ref{fig:timing}), compensated for bedrock deformation, against
        observed trimline elevations for the upper Rhone Glacier
        \citep[Table~1]{Kelly.etal.2004}. Colours
        show the age of maximum ice thickness (as in Fig.~\ref{fig:timing}).
        Model variables were bilinearly interpolated to the trimline locations.
        \textbf{(b)} Histogram of differences between modelled LGM ice surface
        elevation and trimline elevations (50\,m bands). The average
        difference is 861\,m.
        \textbf{(c)} Observed upper Rhone Valley trimline locations
        \citep[Table~1]{Kelly.etal.2004}, modelled age of maximum ice thickness
        (colour) and LGM ice surface elevation (200\,m contours). Hatches mark
        the LGM cold-based areas (basal temperature above $1 \times 10^{-3}$\,K
        below freezing at the age of maximum ice thickness).}
      \label{fig:trimlines}
    \end{figure}

@@

    In the upper Rhone Valley, the maximum ice surface elevation reached
    during MIS~2 in our simulation, corrected for bedrock deformation, is
    consistently modelled to have occured several hundred metres above the
    observed trimline elevations (Fig.~\ref{fig:trimlines}a and b), with a mean
    difference of 861\,m (and a standard deviation 197\,m,
    Fig.~\ref{fig:trimlines}b), which is significant in respect to the modelled
    maximum ice thickness in the Rhone Valley (and the Alps) of 2586\,m. This
    result depends on uncertain basal sliding and ice rheological parameters
    to which the model sensitivity was not tested here. However, a previous
    sensitivity study conducted with PISM and a very similar model set-up on
    the Cordilleran ice sheet show that varying ice rheological parameters
    resulted in relative errors in modelled total ice volume of 25\,\unit{\%}
    with regard to ice rheological parameters and 14\,\unit{\%} regarding basal
    sliding parameters \citep[Fig.~7]{Seguinot.etal.2016}. In addition,
    regional sensitivity tests in the Rhone Valley show that a modelled ice
    surface compatible with the trimline elevations is incompatible with the
    documented overspilling flow to the south across the Simplon pass
    \citep{Becker.etal.2017}.
    \todo{Add \citep{Cohen.etal.2018}, and more?}
